Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-12-19 draft.tex
.

\subsubsection{Distribution of Firing Rates}
Achieving a broad distribution of firing rates among neurons was the core motivation for the implementation of diffusive homeostasis. Figure \ref{Fir_Rate_Dist_Compare} shows a first result, comparing both homeostatic mechanisms. As expected, non-diffusive homeostasis leads to a sharp distribution of firing rates at 3 Hz. Diffusive homeostasis indeed results in a much broader distribution of mean firing rates. A large number of experimental studies have found that distributions of firing rates are not only broadly distributed but well described by a log-normal distribution, which has a non-zero third moment or skewness (\cite{Buzsaki_Fir_Rates_2014}, \cite{Wohrer_Fir_Rates_2012}). By definition, the logarithm of the random variable in question is thus again normally distributed. To check for this property we plotted the the distribution of decadic logarithms of firing rates in Figure \ref{Fir_Rate_Dist_Compare} (B). In (A), we found a skewness of $\mathrm{v_{Diff} = 0.765}$, in (B) $\mathrm{v_{Diff,log} = -0.488}$. Though this told us that the distribution is "more symmetric" on a logarithmic scale, it should rather be seen as being neither strictly normally or log-normally distributed.  
\begin{figure}
%%% plot_frequ_dist_mult_sets.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_e_compare.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_i_compare.png}
\caption{Histograms of mean firing rates over the excitatory/inhibitory population in regular (A/A*) and logarithmic space (B/B*). For diffusive homeostasis ($\mathrm{D=1000\, \mu m^2 s^{-1}}$), the distribution was generated from 10 simulation runs, 1 simulation was used for non-diffusive homeostasis. Mean firing rates where calculated from spikes within $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Compare}
\end{figure}
In contrast, inhibitory cells showed a much clearer distinction: A skewness of $\mathrm{v_{Diff} = 1.522}$ in regular space and $\mathrm{v_{Diff,log} = -0.147}$ in log-space. As a first interpretation, we related this difference to the fact that inhibitory cells are only indirectly affected by diffusive homeostasis in the sense that their distribution of activity is mainly determined by differences withing neural input since inhibitory thresholds are all initialized and kept at the same value.

@@

A constant "rate" of an approximately randomly spiking neuron means that the dependence of fluctuations of rates onto the width of the averaging window should be identical to a homogeneous poisson process of the same total mean firing rate. As bin sizes increase, possible long-term fluctuations of firing rates can then be identified as deviations from the variance one would expect from a homogeneous process. We tested this in Figure \ref{f_Var_vs_n_Bin}. (A) was taken from $\mathrm{1000-1500\,s}$, which is well inside the phase of diffusive homeostasis. As one can see, the statistics of the three representative neurons are well fitted by the predictions for homogeneous poisson processes with the same mean rate, being a straight line given by $\mathrm{var = f_{mean}\cdot n_{bins}/T_{total}}$, though slight deviations can be seen as a different factor of proportionality. (B) acts as an illustration showing a counterexample: here, we included the entire time span of the simulation, including the transition between non-diffusive and diffusive homeostasis. For almost all of the neurons, this will cause a transition of firing rate between 3 Hz and a new rate different from 3 Hz. This inhomogeneity is represented by the increasing positive relative deviation from the dashed line for small number of bins (i.e., large bin widths), even increasing again for a very few number of bins.
\begin{figure}
%%% plot_spt_var_vs_bin_width.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/sim_f_var_vs_hom_poisson.png}
\caption{A: Relation between Variance of the frequency signal (acquired by binning spikes) and the number of bins within the given time range of $\mathrm{1000-1500\,s}$, taken from three randomly picked excitatory neurons. B: Same procedure as in (A) for a randomly chosen excitatory neuron, but time range was $\mathrm{1000-1500\,s}$, which includes the initial phase of non-diffusive homeostasis. In both plots, dashed lines are the expected curve of a homogeneous poisson process of the same mean rate.}
\label{f_Var_vs_n_Bin}
\end{figure}
We therefore concluded that diffusive homeostasis manages to retain firing rates on a constant level, not only population-wise but on a single-cell level as well, making Figure \ref In particular, plastic modifications across the network due to STDP and synaptic growth/pruning (which where always acting for all the simulation results presented so far) appeared to be compensated for.

Sweeney et al. found that diffusive homeostasis maintains broadness of firing rates across a wide range of diffusion constants but rapidly approaching zero for small values \cite[p. 6]{Sweeney_Paper}. We were able to reproduce this result, see Figure \ref{Fir_Rate_Dist_Width_vs_D}. Homeostasis reaches a point of saturation, where faster diffusion has no effect on the heterogeneity of firing rates. 
\begin{figure}
%%% freq_dist_width_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/std_dev_vs_D_neumann.png}
\caption{Standard deviation of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Width_vs_D}
\end{figure}
To further quantify this dependence, we also investigated the influence of the diffusion constant onto the distribution's skewness, shown in Figure \ref{Fir_Rate_Dist_Skew_vs_D}. Compared to the standard deviation, we see a similar but not as clear trend with a drop for very small diffusion constants, even occasionally resulting in a left-skewed distribution (negative D-values).
\begin{figure}
%%% freq_dist_skew_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/skew_vs_D_neumann.png}
\caption{Skewness of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Skew_vs_D}
\end{figure}

!!!HEAD!!!

>>>>

\subsubsection{Distribution of Firing Rates}
Achieving a broad distribution of firing rates among neurons was the core motivation for the implementation of diffusive homeostasis. Figure \ref{Fir_Rate_Dist_Compare} shows a first result, comparing both homeostatic mechanisms. As expected, non-diffusive homeostasis leads to a sharp distribution of firing rates at 3 Hz. Diffusive homeostasis indeed results in a much broader distribution of mean firing rates. A large number of experimental studies have found that distributions of firing rates are not only broadly distributed but well described by a log-normal distribution, which has a non-zero third moment or skewness (\cite{Buzsaki_Fir_Rates_2014}, \cite{Wohrer_Fir_Rates_2012}). By definition, the logarithm of the random variable in question is thus again normally distributed. To check for this property we plotted the the distribution of decadic logarithms of firing rates in Figure \ref{Fir_Rate_Dist_Compare} (B). In (A), we found a skewness of $\mathrm{v_{Diff} = 0.765}$, in (B) $\mathrm{v_{Diff,log} = -0.488}$. Though this told us that the distribution is "more symmetric" on a logarithmic scale, it should rather be seen as being neither strictly normally or log-normally distributed.  
\begin{figure}
%%% plot_frequ_dist_mult_sets.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_e_compare.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/fir_rate_dist_i_compare.png}
\caption{Histograms of mean firing rates over the excitatory/inhibitory population in regular (A/A*) and logarithmic space (B/B*). For diffusive homeostasis ($\mathrm{D=1000\, \mu m^2 s^{-1}}$), the distribution was generated from 10 simulation runs, 1 simulation was used for non-diffusive homeostasis. Mean firing rates where calculated from spikes within $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Compare}
\end{figure}
In contrast, inhibitory cells showed a much clearer distinction: A skewness of $\mathrm{v_{Diff} = 1.522}$ in regular space and $\mathrm{v_{Diff,log} = -0.147}$ in log-space. As a first interpretation, we related this difference to the fact that inhibitory cells are only indirectly affected by diffusive homeostasis in the sense that their distribution of activity is mainly determined by differences withing neural input while inhibitory thresholds are all initialized and kept at the same value.

@@

A constant "rate" of an approximately randomly spiking neuron means that the dependence of fluctuations of rates onto the width of the averaging window should be identical to a homogeneous poisson process of the same total mean firing rate. As bin sizes increase, possible long-term fluctuations of firing rates can then be identified as deviations from the variance one would expect from a homogeneous process. We tested this in Figure \ref{f_Var_vs_n_Bin}. (A) was taken from $\mathrm{1000-1500\,s}$, which is well inside the phase of diffusive homeostasis. As one can see, the statistics of the three representative neurons are well fitted by the predictions for homogeneous poisson processes with the same mean rate, being a straight line given by $\mathrm{var = f_{mean}\cdot n_{bins}/T_{total}}$, though slight deviations can be seen as a different factor of proportionality. (B) acts as an illustration showing a counterexample: here, we included the entire time span of the simulation, including the transition between non-diffusive and diffusive homeostasis. For almost all of the neurons, this will cause a transition of firing rate between 3 Hz and a new rate different from 3 Hz. This inhomogeneity is represented by the increasing positive relative deviation from the dashed line for small number of bins (i.e., large bin widths), even increasing again for a very few number of bins.
\begin{figure}
%%% plot_spt_var_vs_bin_width.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/sim_f_var_vs_hom_poisson.png}
\caption{A: Relation between Variance of the frequency signal (acquired by binning spikes) and the number of bins within the given time range of $\mathrm{1000-1500\,s}$, taken from three randomly picked excitatory neurons. B: Same procedure as in (A) for a randomly chosen excitatory neuron, but time range was $\mathrm{1000-1500\,s}$, which includes the initial phase of non-diffusive homeostasis. In both plots, dashed lines are the expected curve of a homogeneous poisson process of the same mean rate.}
\label{f_Var_vs_n_Bin}
\end{figure}
We therefore concluded that diffusive homeostasis manages to retain firing rates on a constant level, not only population-wise but on a single-cell level as well. In particular, plastic modifications across the network due to STDP and synaptic growth/pruning - which where always acting for all the simulation results presented so far - appeared to be compensated for.

Sweeney et al. found that diffusive homeostasis maintains broadness of firing rates across a wide range of diffusion constants but rapidly approaching zero for small values \cite[p. 6]{Sweeney_Paper}. We were able to reproduce this result, see Figure \ref{Fir_Rate_Dist_Width_vs_D}. Homeostasis reaches a point of saturation, where faster diffusion has no effect on the heterogeneity of firing rates. 
\begin{figure}
%%% freq_dist_width_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/std_dev_vs_D_neumann.png}
\caption{Standard deviation of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Width_vs_D}
\end{figure}
To further quantify this dependence, we also investigated the influence of the diffusion constant onto the distribution's skewness, shown in Figure \ref{Fir_Rate_Dist_Skew_vs_D}. Compared to the standard deviation, we see a similar but not as clear trend with a drop for very small diffusion constants, even occasionally resulting in a left-skewed distribution (negative D-values).
\begin{figure}
%%% freq_dist_skew_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/skew_vs_D_neumann.png}
\caption{Skewness of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Skew_vs_D}
\end{figure}
A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, simulation-wise this case is quite easy to simulate: one simply has  to feed all NO-sources into a single scalar variable of NO concentration. In particular, this will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time. Since we initialized all excitatory neurons at the same threshold, only one global excitatory threshold variable had to be considered in this case. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory was $\mathrm{\sigma_{inst} = 1.41\, Hz}$ and the skewness $\mathrm{v_{inst} = 1.27}$ ($\mathrm{v_{inst,log} = -0.46}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 2 simulations and $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

Summing up the results of this section, we can state that stable network activity can be achieved in the SORN with diffusive homeostasis while producing a broad distribution of firing rates within the excitatory population. We found that this ensemble of activity is maintained on a single cell level, each cell spiking with approximately poissonian statistics at a constant mean rate. The broadness of distribution could be maintained at a relatively constant level over a wide range of diffusion rates. However, we did not find a distribution as right skewed and heavy tailed as reported by Sweeney et al. and numerous experimental studies.

\subsection{Network Topology}
So far, we have only analyzed network activity, whereas in this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the main subject of our investigations. Moreover, in contrast to the previous section, we did not explicitly aim for the emergence of an explicitly \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the SORN.
